---
title: "Comparison of Differnetial Expression Analysis on Simulated Data with DESeq2"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{vignette_1123}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Here, we compare our simpleDE method with DESeq2, one of the most popular R packges for diffrential expression analyis. Both methods will be performed on the same simulated dataset, and compared in terms of correctness and effciency.

```{r setup}
options(warnings=-1)
library(simpleDE)
suppressPackageStartupMessages(library(DESeq2))

```

Here we simulate a dataset containing 10000 genes expression data in 10 samples from two condition (5 treatment vs 5 control). Among those 10000 genes, 100 genes are differentially expressed genes, simulated from two negative binomial distribution with distinct means. 

```{r}
set.seed(20191124)
mu = rpois(10000, 20)
size = 20
ctr_dt = matrix(NA, 10000, 5)
trt_dt = matrix(NA, 10000, 5)
for (i in 1:100){
  ctr_dt[i, ] = rnbinom(n = 5, size =  size, mu = mu[i])
  trt_dt[i, ] = rnbinom(n = 5, size =  size, mu = mu[i] + 40)
}
for (i in 101:10000){
  ctr_dt[i, ] = rnbinom(n = 5, size = size, mu = mu[i])
  trt_dt[i, ] = rnbinom(n = 5, size = size, mu = mu[i])
}

dt = cbind(ctr_dt, trt_dt)
colnames(dt) = c(sapply(1:5, function(x) paste0("ctr",x)), sapply(1:5, function(x) paste0("trt",x)))
rownames(dt) = c(sapply(1:100, function(x) paste0("DEG",x)), sapply(1:9900, function(x) paste0("non_DEG",x)))

dt[1:10,]
dt[101:110,]

```

Here we also set up the sample information and gene information.

```{r}
# Create simpleDE
mtx = dt
s_data = data.frame(treatment = rep(c(0,1), each = 5))
g_data = data.frame(is_DEG = c(rep(1, 100), rep(0, 9900)))
```

Now we perform simpleDE and get the final differnetially expressed genes by threhold of 0.5 for log fold-change and 0.01 for p value.
```{r}
simpleDE.test = function(){
  simple.dt = Create_simpleDE(mtx, s_data, g_data)
  simple.dt = Filter_Data(simple.dt, verbose = F)
  simple.dt = Normalize_Data(simple.dt, option = "size_factor")
  de.geno = DEA_test(simple.dt, group = "treatment", verbose = F)
  de.res = Get_DEG(de.geno, "log_FC", 0.5, is.sort = F)
  de.res = Get_DEG(de.res, "p_value", 0.01, is.sort = F)
  print(head(de.res))
  print(table(substr(rownames(de.res), 1, 3)))
}
# simpleDE.test()
system.time(simpleDE.test())

```
Now we also perform DESeq2 on the same data.
```{r}
DESeq2.test = function(){
  dds = DESeqDataSetFromMatrix(countData = dt, colData = s_data, design = ~ treatment)
  dds = DESeq(dds)
  res = results(dds)
  res = res[ res$log2FoldChange > log(exp(0.5)) | res$log2FoldChange < -log(exp(0.5)), ]
  res = res[res$pvalue < 0.01, ]
  print(head(res))
  print(table(substr(rownames(res), 1, 3)))
}
# DESeq2.test()
system.time(DESeq2.test())
```

# Conclusion

In terms of sensitivity, while DESeq2 has a full score, simpleDE has a decent 87% true positive rate.
In terms of specificity, DESeq2 contains a level of (9900-54) / 9900 = 99.45% and simpleDE has a level of (9900 - 33) / 9900 = 99.67%.
Overall, simpleDE is a less conservative method comparing with DESeq2.

In terms of time effiency, simpleDE is much faster than DESeq2 due to simplicity of algorithm and less handling of data structure.
